---
title: "Introduction to Tidyverse"
tutorial:
  id: "cp_bootcamp_2021_chp4_practice"
  version: 1.0
output: 
  learnr::tutorial:
    theme: "lumen"
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn the tidyverse ecosystem and use this paradigm to explore data
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = "___+")

library(gapminder)
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
data("mtcars")
```


## Prep

We'll first load the gampinder data and then fix it so that the `country` and `continent` columns are text instead of 'factors'
```{r}
library(gapminder)
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
```

For this practice, try to use the pipe (`%>%`) wherever possible to chain together operations, with the general pattern

```
input_data %>% 
  operation_1 %>% 
  operation_2
```

If you want to capture the result of the chain of operations and assign it to a new variable (rather than say printing out the result to the console), you can do:

```
result <- input_data %>% 
  operation_1 %>% 
  ...
```

## Filter 

Recall that the `filter` function is useful for extracting a subset of rows from a data table. Each input to `filter` (in addition to the data table itself) specifies a logical condition that will be applied to select rows. If you provide multiple inputs `filter(data, A, B)` this selects for 'A AND B'. You can use `A | B` to specify 'A OR B'.

Let's practice some filtering. Use the `filter` function (and the `%>%`) to extract the specified set of rows from the `gapminder` table.

The first 5 rows where continent is equal to `Asia` and life expectancy is greater than 35 yrs
```{r filter1, exercise = TRUE}

```

```{r filter1-solution}
gapminder %>%
  filter(continent == "Asia", lifeExp > 35) %>%
  head(5)
```


All rows where the country begins with 'A' 
(hint: the `>` and `<` operators also work on strings to compare alphabetical order)
```{r filter2, exercise = TRUE}

```

```{r filter2-solution}
gapminder %>%
  filter(lifeExp < 'B')
```


All rows where the continent is `Oceania` OR the country is one of Turkey, Uruguay, or Taiwan
```{r filter3, exercise = TRUE}

```

```{r filter3-solution}
# gapminder %>% 
#   filter(continent == "Oceania" | country %in% c('Turkey', 'Uruguay', 'Taiwan'))
```



## Arrange/Select

Recall that `arrange` sorts the rows of a data table based on the values in one or more columns. 

`select` allows you to select a subset of columns from a table as: `select(col1, col2, col3)`.

Sort the rows of the `gapminder` table alphabetically using the continent and country columns. First by continent, then by country
```{r arrange1, exercise = TRUE}

```

```{r arrange1-solution}
gapminder %>%
  arrange(continent, country)
```


What were the top 5 years with the highest life expectancy in China? Use the `select` function to create a table that contains only the year and lifeExp for these top 5 entries.
```{r arrange2, exercise = TRUE}

```

```{r arrange2-solution}
gapminder %>%
  filter(country == "China") %>%
  arrange(desc(lifeExp)) %>%
  head(5) %>%
  select(year,lifeExp)
```



## Mutate

Recall that `mutate` allows you to add new columns to a table. The new columns can also be generated as a function of existing columns in the table (e.g. `in_data %>% mutate(col12_sum = old_col1 + old_col2)`).

Add a new variable to the gapminder table called `GDP` which has the total GDP for each country
```{r mutate1, exercise = TRUE}

```

```{r mutate1-solution}
gapminder <- gapminder %>%
  mutate(GDP = pop * gdpPercap)
```


```{r gdp_setup, include = FALSE, echo = FALSE}
#adding GDP to gapminder so the rest will work
gapminder <- gapminder %>%
  mutate(GDP = lifeExp * gdpPercap)
```

```{r, ref.label="gdp_setup", echo = FALSE}
```

Add a column called `high_GDP` to gapminder which has value TRUE if the `GDP` is higher than the overall average, and FALSE otherwise. (It's easiest to do this in two steps).

```{r mutate2, exercise = TRUE, exercise.setup='gdp_setup'}

```

```{r mutate2-solution}
mean_gdp <- mean(gapminder$GDP) #first calculate the overall avg GDP
gapminder %>%
  mutate(high_GDP = GDP > mean_gdp)
```



## Group-by/summarise

Recall that `summarise` allows you to compute summary statistics across rows of a table. Use `summarise` to compute the mean and median GDP for counties in Asia after 1980. Your result should be a table with a single row and two columns called `mean_gdp` and `median_gdp`

```{r summarise1, exercise = TRUE, exercise.setup='gdp_setup'}

```

```{r summarise1-solution}
gapminder %>%
  filter(continent == 'Asia', year >= 1980) %>%
  summarise(mean_gdp = mean(GDP),
            median_gdp = median(GDP))
```


Recall that `group_by` allows you to group rows by a variable or combinations of variables, which is typically used in combination with `summarise` to compute summary statistics for each group in a table.

Make a table with the mean, median, max, and min life expectancies for Asian countries in the 1970's.
```{r group_by1, exercise = TRUE, exercise.setup='gdp_setup'}

```

```{r group_by1-solution}
gapminder %>%
  filter(continent == "Asia", year>=1970, year<1980) %>%
  group_by(country) %>%
  summarise(mean_life_exp = mean(lifeExp),
            median_life_exp = median(lifeExp),
            max_life_exp = max(lifeExp),
            min_life_exp = min(lifeExp))
```


Make a table with the mean lifeExp by country and year
```{r group_by2, exercise = TRUE, exercise.setup='gdp_setup'}

```

```{r group_by2-solution}
gapminder %>%
  group_by(country, year) %>%
  summarise(mean_life_exp = mean(lifeExp))
```


*Challenge*: Make a table with the mean lifeExp by country and decade. (Hint: you can use the `floor` function to round numbers down to an integer)
```{r group_by3, exercise = TRUE, exercise.setup='gdp_setup'}
# # The floor function will always round down. So when we divide the year by 10 , floor and then multiply back, this is equavalent to stripping it's unit digit. This gives the decade

```

```{r group_by3-solution}
gapminder %>%
  mutate(decade = floor(year/10)*10) %>%
  group_by(country, decade) %>%
  summarise(mean_life_exp = mean(lifeExp))
```


## More practice

### more with gapminder

Find the five entries in gapminder that had the highest ratio of GDP to life expectancy, using only countries in Asia for years after 1980
```{r comb_practice1, exercise = TRUE, exercise.setup='gdp_setup'}

```

```{r comb_practice1-solution}
gapminder %>%
  mutate(gpd_to_lifeexp = GDP/lifeExp) %>%
  filter(continent == "Asia", year > 1980) %>%
  arrange(desc(gpd_to_lifeexp)) %>%
  head(5)
```


*Challenge*: Make a table which lists the number of samples where each continent had a mean lifeExpectancy greater than 60. 
```{r group_by4, exercise = TRUE, exercise.setup='gdp_setup'}

```

```{r group_by4-solution}
gapminder %>%
  group_by(continent, year) %>%
  summarise(mean_life_exp = mean(lifeExp)) %>%
  filter(mean_life_exp > 60) %>%
  group_by(continent) %>%
  summarise(number_of_samples = n())
```



### mtcars

Some additional practice (optional) using the `mtcars` dataset which contains info about cars. This practice is taken from the HBC course [here](https://hbctraining.github.io/Training-modules/Tidyverse_ggplot2/lessons/tidyverse_exercise.html)

```{r}
head(mtcars)
```


```{r mtcar_setup, echo = FALSE}
# Turn row name to column and convert to tibble
mtcars_tb <- mtcars %>% 
  rownames_to_column(var = "car") %>% 
  as_tibble()
```

```{r, ref.label="mtcar_setup", echo = FALSE}
```

Perform the following data wrangling steps using `mtcars_tb` as the starting point. Name the final variable as mtcars_final.

- One of the columns is `am`. It indicates transmission status, where 0 refers to automatic, and 1 refers to manual. Extract cars with manual transmission status.

- We are only interested in these five columns: `car`, `mpg`, `cyl`, `wt`, `am`. Select only these five columns for further analysis.

- Some column names are not intuitive. Rename the `cyl` to cylinder, `wt` to weight, and `am` to transmission.

- We want to order our data. Arrange the data first by cylinder in ascending order, and then by mpg in descending order.

Check your result: if you finish the data wrangling successfully, the mtcars_final should have 13 entries with 5 features. The first entry should be Toyota Corolla, and the last entry should be Maserati Bora.

```{r mtcars_practice, exercise = TRUE, exercise.setup='mtcar_setup'}

```

```{r mtcars_practice-solution}
mtcars_final <- mtcars_tb %>%
  select(car, mpg, cylinder = cyl, weight = wt, transmission = am) %>%   filter(transmission == 1) %>%
  arrange(cylinder, desc(mpg))
mtcars_final
```


