---
title: "Data Vizualization"
tutorial:
  id: "cp_bootcamp_2021_chp5_practice"
  version: 1.0
output: 
  learnr::tutorial:
    theme: "lumen"
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to make beautiful publication ready plots
---

```{r setup, include=FALSE}
library(learnr)
library(gapminder)
library(ggrepel)
library(tidyverse)
library(plotly)


knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 5, fig.height = 3.5)
tutorial_options(exercise.timelimit = 20, exercise.blanks = "___+")


gapminder <- gapminder %>% 
  mutate(country = as.character(country),
         continent = as.character(continent))

gap_92 <- gapminder %>% 
  filter(year == 1992) %>% 
  mutate(gdp = pop*gdpPercap/1e9)
```


## Scatterplot basics {data-progressive=TRUE}

### Create data

We'll be using the gapminder data again

```{r gapminder_head}
head(gapminder)
```

First, pull out the gapminder data from the year 1992, and add a column `gdp` (in units of billions of $). Call this new table `gap_92`

```{r make_gap_df, exercise = TRUE}
gap_92 <- ___
```

```{r make_gap_df-solution}
gap_92 <- gapminder %>% 
  filter(year == 1992) %>% 
  mutate(gdp = pop*gdpPercap/1e9)
```

### Recreate Graph 1

For the following you will be able to use the `gap_92` table you created above.
Let's make a basic scatterplot of the data using `ggplot`. 

Your output should look like this.

```{r, echo = FALSE}
ggplot(data = gap_92, mapping = aes(x = gdp, y = lifeExp)) + 
  geom_point()
```


```{r gap92_setup}
gap_92 <- gapminder %>% 
  filter(year == 1992) %>% 
  mutate(gdp = pop*gdpPercap/1e9)
```

```{r scatter_test1, exercise = TRUE,  exercise.setup='gap92_setup'}

```

```{r scatter_test1-solution}
ggplot(data = gap_92, mapping = aes(x = gdp, y = lifeExp)) + 
  geom_point()
```




### Recreate Graph 2

Now make the same figure, but set the x-axis on a log10 scale, add informative x- and y-axis labels, and a figure title. 

Your new plot should look like this:

```{r, echo=F}
ggplot(data = gap_92, mapping = aes(x = gdp, y = lifeExp)) + 
  geom_point() +
  scale_x_log10() +
  labs(x = 'GDP (billions $)', y = 'Life expectancy', title = 'Life Expectancy vs GDP')
```

```{r scatter_test2, exercise = TRUE,  exercise.setup='gap92_setup'}

```

```{r scatter_test2-solution}
ggplot(data = gap_92, mapping = aes(x = gdp, y = lifeExp)) + 
  geom_point() +
  scale_x_log10() +
  labs(x = 'GDP (billions $)', y = 'Life expectancy', title = 'Life Expectancy vs GDP')
```



### Additional Practice

Now explore mapping additional information from this dataset, such as `continent` and `pop` to different aesthetic variables of the points. Use `help(geom_point)` to see information on the available aesthetics for points.
```{r scatter_explore, exercise = TRUE}

```



## Line plots and color basics

### Line plot base


Pick 3 countries and make a plot of their lifeExp over time including both lines and points. Make the lines colored by country, but make the points all black. 

Your plot should look similar to this.

```{r, echo = FALSE}
target_countries <- c('Venezuela', 'Vietnam', 'Uruguay')
gap_targ <- gapminder %>% 
  filter(country %in% target_countries)
ggplot(gap_targ, mapping = aes(x = year, lifeExp, color = country)) + 
  geom_line() + 
  geom_point(color = 'black')
```


```{r lines_and_colors1, exercise = TRUE}

```

```{r lines_and_colors1-solution}
## One of the solution
target_countries <- c('Venezuela', 'Vietnam', 'Uruguay')
gap_targ <- gapminder %>% 
  filter(country %in% target_countries)
ggplot(gap_targ, mapping = aes(x = year, lifeExp, color = country)) + 
  geom_line() + 
  geom_point(color = 'black')
```

### Transparency and size

Try controlling the size and transparency of the points. (transparancy is controlled by 'alpha', which ranges from 0 to 1)

```{r lines_and_colors2, exercise = TRUE}

```

```{r lines_and_colors2-solution}
## One of the solution
target_countries <- c('Venezuela', 'Vietnam', 'Uruguay')
gap_targ <- gapminder %>% 
  filter(country %in% target_countries)
ggplot(gap_targ, mapping = aes(x = year, lifeExp, color = country)) + 
  geom_line() + 
  geom_point(color = 'black', size=3, alpha=0.4)
```


Use `scale_color_manual` to set the colors to ones you like. You can choose specific colors for each country using a named vector of colors like this: `my_colors = c(country1 = 'darkred', country2 = 'magenta', ...)`. You can see names of available colors [here](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf). 

```{r lines_and_colors3, exercise = TRUE}

```

```{r lines_and_colors3-solution}

my_colors = c(Venezuela = 'darkred', Vietnam = 'magenta', Uruguay='blue')

target_countries <- c('Venezuela', 'Vietnam', 'Uruguay')
gap_targ <- gapminder %>% 
  filter(country %in% target_countries)
ggplot(gap_targ, mapping = aes(x = year, lifeExp, color = country)) + 
  geom_line() + 
  geom_point(color = 'black', size=3, alpha=0.4)+
  scale_color_manual(values = my_colors)
```


Now try picking one of the default RColorBrewer themes, using the `scale_color_brewer` function. See available palettes [here](http://applied-r.com/rcolorbrewer-palettes/)

```{r lines_and_colors4, exercise = TRUE}

```

```{r lines_and_colors4-solution}
target_countries <- c('Venezuela', 'Vietnam', 'Uruguay')
gap_targ <- gapminder %>% 
  filter(country %in% target_countries)
ggplot(gap_targ, mapping = aes(x = year, lifeExp, color = country)) + 
  geom_line() + 
  geom_point(color = 'black', size=3, alpha=0.4)+
scale_colour_brewer(palette = "Set1")
```



## geom_smooth

Make a plot using `geom_smooth` to show the average trends of lifeExp over time for each continent. Test out using both linear regression (method = 'lm') and 'loess' models (the default). Also try out turning off the confidence intervals (parameter `se` in `geom_smooth`). 

For reference, your plots should look something like this:

```{r, echo = FALSE}
ggplot(gapminder, aes(year, lifeExp, color = continent)) + 
  geom_smooth(se = FALSE)
```

```{r geom_smooth, exercise = TRUE}

```

```{r geom_smooth-solution}
ggplot(gapminder, aes(year, lifeExp, color = continent)) + 
  geom_smooth(se = FALSE)
```



## geom_facet

Using only data from up to and including the year 1962, make a faceted scatterplot of gdpPercap vs lifeExp, where there is a separate scatterplot for each year (use `facet_wrap()`).
Make gdpPercap on a log10 scale
Add a 'smoothing trend' with `geom_smooth()` if you want!

It should look like this:

```{r, echo = FALSE}
gap_early <- gapminder %>% filter(year <= 1962)
ggplot(gap_early, mapping = aes(gdpPercap, lifeExp)) + 
  scale_x_log10() +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_wrap(~year)
```


```{r geom_facet, exercise = TRUE}

```

```{r geom_facet-solution}
gap_early <- gapminder %>% filter(year <= 1962)
ggplot(gap_early, mapping = aes(gdpPercap, lifeExp)) + 
  scale_x_log10() +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_wrap(~year)
```


## Adding text 

Make a table called `continent_stats` which has the average lifeExp and average gdpPercap for each continent. Then use to it make a scatterplot of avg lifeExp vs avg gdp, and label the points using `geom_label_repel()` (or `geom_text_repel()` if you prefer)

Your plot should look like this:

```{r, echo = FALSE}
continent_stats <- gapminder %>% 
  group_by(continent) %>% 
  summarise(avg_lifeExp = mean(lifeExp),
            avg_gdpPercap = mean(gdpPercap))
ggplot(continent_stats, mapping = aes(avg_lifeExp, avg_gdpPercap, label = continent)) + 
  geom_point() + 
  geom_label_repel()
```

```{r adding_text, exercise = TRUE}

```

```{r adding_text-solution}
continent_stats <- gapminder %>% 
  group_by(continent) %>% 
  summarise(avg_lifeExp = mean(lifeExp),
            avg_gdpPercap = mean(gdpPercap))
ggplot(continent_stats, mapping = aes(avg_lifeExp, avg_gdpPercap, label = continent)) + 
  geom_point() + 
  geom_label_repel()
```



## Plotting distributions

### Violin plots
Make a violin plot of lifeExp for each continent 
```{r violin1, exercise = TRUE}

```

```{r violin1-solution}
ggplot(gapminder, aes(x=continent, y=lifeExp))+
  geom_violin()
```


Now make each continent's violin a different fill color (play around with using `fill` vs `color` to encode the continent)

Try the fill aesthetic

```{r violin2, exercise = TRUE}

```

```{r violin2-solution}
ggplot(gapminder, aes(x=continent, y=lifeExp, fill=continent))+
  geom_violin()
```

Try the color aesthetic

```{r violin3, exercise = TRUE}

```

```{r violin3-solution}
ggplot(gapminder, aes(x=continent, y=lifeExp, color=continent))+
  geom_violin()
```

### Boxplot plots

Try it as a boxplot
(Make a boxplot plot of lifeExp for each continent )

```{r boxplot1, exercise = TRUE}

```

```{r boxplot1-solution}
ggplot(gapminder, aes(x=continent, y=lifeExp))+
  geom_boxplot()
```

Use `fct_reorder` to reorder the continents by median lifeExp before making your boxplot. The inputs to `fct_reorder` are a categorical variable, a value variable to order the categories by, and a summary function. So, `fct_reorder(country, gdp, max)` would reorder the country categorical variable based on max GDP across years.

```{r boxplot_reorder, exercise = TRUE}

```

```{r boxplot_reorder-solution}
gapminder <- gapminder %>% mutate(continent = fct_reorder(continent, lifeExp, median))
ggplot(gapminder, aes(x=continent, y=lifeExp))+
  geom_boxplot()
```

Try making a density plot with `geom_density()`, separate continents by fill color

```{r density_plot, exercise = TRUE}

```

```{r density_plot-solution}
ggplot(gapminder, aes(fill=continent, x=lifeExp))+
  geom_density()
```

What happens when you make the density plots semi-transparent (i.e. setting `alpha` to ~0.5)?
Try the alpha property to change transparency

```{r density_plot_alpha, exercise = TRUE}

```

```{r density_plot_alpha-solution}
ggplot(gapminder, aes(fill=continent, x=lifeExp))+
  geom_density(alpha=0.5)
```

## Using theme

### Rotate axis

Now let's make a bar plot of the gdpPercap for countries in the Americas in the year 1982. Recall that `geom_col` will do this. Your plot should initially look like this:

```{r gap82_setup}
gap_82 <- gapminder %>% 
  filter(year == 1982, continent == 'Americas') 
```

```{r, echo = FALSE}
gap_82 <- gapminder %>% 
  filter(year == 1982, continent == 'Americas') 
ggplot(gap_82, aes(x = country, y = gdpPercap)) + 
  geom_col()
```

```{r plot_count, exercise = TRUE, exercise.setup='gap82_setup'}
ggplot(gap_82, )
```

Let's try to figure out how to fix the x-axis labels so they are vertical (and not overlapping). See if you can use Google to figure out how to rotate the x-axis labels in ggplot to make them vertical.

```{r rotate_labels, exercise = TRUE, exercise.setup='gap82_setup'}
ggplot(gap_82, aes(x = country, y = gdpPercap)) + 
  geom_col() +
  ___
```

```{r rotate_labels-solution}
ggplot(gap_82, aes(x = country, y = gdpPercap)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

### Reorder countries

Then see if you can reorder the countries from highest to lowest. Here again you can use the `fct_reorder` function, inside a `mutate` to order countries by gdpPercap. Note that there's another optional input to `fct_reorder` called `.desc` which is FALSE by default 

Your plot should look like this
```{r, echo = FALSE}
gap_82 <- gap_82 %>% 
  mutate(country = fct_reorder(country, gdpPercap, .desc = TRUE))
ggplot(gap_82, aes(x = country, y = gdpPercap)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


```{r rotate_labels_ordered, exercise = TRUE, exercise.setup='gap82_setup'}

```

```{r rotate_labels_ordered-solution}
gap_82 <- gap_82 %>% 
  mutate(country = fct_reorder(country, gdpPercap, .desc = TRUE))
ggplot(gap_82, aes(x = country, y = gdpPercap)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


## plotly

Now take the code from the last example in 'Scatterplot basics' to make a scatterplot of gdp vs life expectancy from countries in 1992, and let's make it an interactive scatterplot using the package `plotly`. All you have to do is save your plot to a variable (say `my_plot`, and then pass it as an input to the function `ggplotly`). Try it out! Try adding an aesthetic `text` (say to encode the country) and see what happens. Your plot should look like this: 

```{r, echo = FALSE}
my_data <- gapminder %>% 
  filter(year == 1992) %>% 
  mutate(gdp = gdpPercap * pop / 1e9) 
my_plot <- ggplot(my_data, aes(gdp, lifeExp, text = country)) + 
  geom_point(aes(color = continent, size = pop)) + 
  scale_x_log10() +
  guides(size = 'none') +
  xlab('Gross Domestic Product (Billions $)') +
  ylab('Life Expectancy at birth (years)')
ggplotly(my_plot)
```


```{r plotly, exercise = TRUE}

```

```{r plotly-hint}
## Adding the text as an aesthetic
ggplot(my_data, aes(text = country))
```


```{r plotly-solution}
my_data <- gapminder %>% 
  filter(year == 1992) %>% 
  mutate(gdp = gdpPercap * pop / 1e9) 
my_plot <- ggplot(my_data, aes(gdp, lifeExp, text = country)) + 
  geom_point(aes(color = continent, size = pop)) + 
  scale_x_log10() +
  guides(size = 'none') +
  xlab('Gross Domestic Product (Billions $)') +
  ylab('Life Expectancy at birth (years)')
ggplotly(my_plot)
```



## Extra practice: GO analysis

For extra practice, try following along with the ggplot tutorial from HBC [here](https://hbctraining.github.io/Training-modules/Tidyverse_ggplot2/lessons/03_ggplot2.html)

The top 30 enriched GO biological process gene results are loaded in the table `bp_oe`
```{r GO_setup, echo = FALSE}
load(here::here('data','GOs_for_plotting.RData'))
bp_oe <- bp_oe %>% head(30)
```

```{r, ref.label='GO_setup'}

```

```{r}
head(bp_oe)
```



The tutorial will walk through how to make a plot that looks like this:
```{r, echo = FALSE}
library(RColorBrewer)
mypalette <- brewer.pal(3, "YlOrRd")

ggplot(bp_oe) +
  geom_point(aes(x = gene_ratio, y = GO_term, color = -log10(p.value)), 
             size = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(size=rel(1.15)),
        axis.text.y = element_text(size=rel(0.7)),
        axis.title = element_text(size=rel(1.15))) +
  xlab("Gene ratios") +
  ylab("Top 30 significant GO terms") +
  ggtitle("Dotplot of top 30 significant GO terms") +
  theme(plot.title = element_text(hjust=0.5, 
  	face = "bold")) +
  scale_color_gradientn(name = "Significance \n (-log10(padj))", colors = mypalette) +
  theme(legend.title = element_text(size=rel(1.15),
	hjust=0.5, 
	face="bold"))
```

You can use this code chunk to practice the different components of the tutorial (you can also do this in a Rmarkdown document if you prefer!)

```{r GO_bp, exercise = TRUE}

```
